Claret: Using Data Types for Highly Concurrent Distributed Transactions

### 2015-02-14 rmd/paper.Rmd
cutting with great abandon

Abstract data types (ADTs) are the solution to this problem. Rather than limiting the records in databases to being primitive types like strings or integers, raising them to more complex data types provides a richer interface, exposing ample opportunities for optimization to the database and a precise mechanism to express the intentions of programmers.

Performance benefits come from understanding the properties of ADT operations: those that commute with each other can be performed concurrently, even on multiple copies of the record. This means that transactions whose operations commute abort less, approaching the performance without transactions. This cleanly captures the cases described above where conflicts were unobservable or ordering didn't matter but in a safe way because any operations that don't in fact commute will be handled with traditional concurrency control. Using insights from multi-core concurrent data structures, we show in \autoref{comm} that it is practical to reason about the matrix of commutativity among operations and build implementations that make the right tradeoff between the amount of concurrency allowed and the efficiency of tracking it.

Selecting the right data type for the job gives programmers a clean, precise way to express their desired behavior. For instance, rather than using a generic integer to generate unique identifiers, a `UUID` type, realizing that contiguity of ids isn't necessary, can be trivially parallelized and distributed. Though this is an extremely simple case (and nearly universally adopted optimization), it fits the same mold as more nuanced decisions, such as choosing to represent the number of times a given post has been "retweeted" as a `HyperLogLog`, which can efficiently yield the approximate number of unique users, rather than a costly precise `Set`. Though selecting data structures for the job at hand is nothing new to programmers, only a handful of databases, such as Redis, MongoDB or Riak, support this flexibility, and they do not use the abstraction it affords to enforce strongly consistent transactions.

@@

Commutativity is well known, especially in distributed systems, for enabling important optimizations. Since the 80s, commutativity has been exploited by database systems designers \cite{Weihl:1988,Fekete:90}, within the safe confines of relational models, where knowledge of query plans and complete control of the data structures allows systems to determine when transactions may conflict. Recently, commutativity has seen a resurgence in systems without a predefined data model, such as NoSQL databases and transactional memory.

In the realm of eventual consistency, commutativity has been leveraged for convergence guarantees.
RedBlue consistency allows executing commutative ("blue") operations locally, knowing they will eventually converge. Similarly, conflict-free replicated data types (CRDTs) \cite{Shapiro:SSS11} define commutative merge functions for all operations to ensure that replicas will converge.

Lynx \cite{Zhang:SOSP13} uses knowledge of some commutative operations to make tracking serializability cheaper in chained transactions. Doppel \cite{Narula:OSDI14} added several explicitly commutative operations on records which they exploited to better handle common high-contention situations such as counters and "top-k lists" in the context of a single node multicore database. Finally, HyFlow \cite{Kim:EuroPar13}, a distributed transactional memory framework, reorders commutative operations on specific data types to execute before others to allow them to operate concurrently on a single version of a record.

@@

Though *commutativity* is often discussed in terms of an operation commuting with all other operations, it is actually more nuanced. If a pair of operations commute, then executing them in either order will produce the same result. Using the definitions from \cite{Kulkarni:PLDI11}, whether or not a pair of method invocations commute is a property of the data type and is a function of the methods, their arguments, their return values, and the *abstract* state of their target. We call the full set of commutativity rules for a data type its *commutativity specification.* An example specification for a *Set* is shown in \autoref{spec}. It is important to note that interface choice affects commutativity: if `add(x)` returned a boolean that expressed if the item was added or not, then `add(x)` would only commute with `add(y)` if $x \ne y$. In this case, the difference doesn't drastically impact commutativity, but it would affect the cost of checking commutativity dynamically.

For a given data type interface, there is one specification that expresses the maximum concurrency, like the one in \autoref{spec}, but there may be many that express varying degrees of commutativity. The space of possible specifications forms a *lattice* where $\top$ specifies the maximum possible commutativity, and $\bot$ represents the case where no operations are allowed to commute. As Kulkarni \cite{Kulkarni:PLDI11} explored, there is a complicated tradeoff between exposing more concurrency and efficiently tracking commutativity. Data structure designers and users navigate these choices to optimize their programs.

@@

Sometimes, due to heavily skewed workloads such as those arising from social networks, there may be so many requests to a single record that just executing all the operations, even without aborting, is a bottleneck which prevents scaling. The Ellen Degeneres selfie retweet is a prime example of this. In addition to transactional boosting, which we evaluate in this work, there are many other uses for commutativity. We mention just a few here which we have yet to evaluate.

### Record splitting for parallelism

Doppel \cite{Narula:OSDI14} observed that several contentious operations, such as incrementing counters or keeping track of the maximum bid, actually commute because the application doesn't need the output of each update operation. This allows them to *split* hot records onto multiple cores and execute commutative operations in parallel, *reconciling* them back to a single value before executing reads or non-commuting operations. This observation can be generalized via ADTs: operations can operate on copies of a record in parallel provided they commute with each other and provided they can be combined at the end of a phase before beginning to execute a new phase with a different set of operations that didn't commute with the first set.

### Combining

Another way to reduce contention on a shared data structure is to synchronize hierarchically: first with a couple immediate neighbors, then possibly with more clusters of neighbors, and finally with the data structure itself. This is known as combining \cite{flatCombining,yew:combining-trees,funnels} in the multi-processing literature, and could be applied to contended records in our model just as well as to shared data structures where it originated.

# Evaluation {#eval}
To demonstrate the efficacy of leveraging commutative operations in transactions, we built a simple prototype key-value store, modeled after Redis, that supports complex data types for records, each with their own set of operations. 
The transaction protocol employs standard two-phase commit, and two-phase locking, more details can be found in the appendix (\autoref{apx:protocol}).

>>>>

Abstract data types (ADTs) are the solution to this problem. Rather than limiting the records in databases to being primitive types like strings or integers, raising them to more complex data types provides a richer interface, exposing ample opportunities for optimization to the database and a precise mechanism to express the intentions of programmers. Understanding the properties of ADT operations, such as which operations commute, can be used in a variety of ways to improve database performance, such as reducing the number of aborts. By choosing the right data type, programmers make tradeoffs between performance and precision based on their application's needs.

@@

Commutativity is well known, especially in distributed systems, for enabling important optimizations. Since the 80s, commutativity has been exploited by database systems designers \cite{Weihl:1988,Fekete:90}, within the safe confines of relational models, where knowledge of query plans and complete control of the data structures allows systems to determine when transactions may conflict. Recently, commutativity has seen a resurgence in systems without a predefined data model, such as NoSQL databases and transactional memory.
Commutativity is leveraged in eventually consistent databases for convergence in work such as RedBlue consistency \cite{Li:OSDI12} and conflict-free replicated data types (CRDTs) \cite{Shapiro:SSS11}. Other systems specialize for commutative operations to improve transaction processing, such as Lynx \cite{Zhang:SOSP13} for tracking serializability, and Doppel \cite{Narula:OSDI14} for executing operations in parallel on highly contended records, and HyFlow \cite{Kim:EuroPar} for reordering operations in the context of distributed transactional memory. We propose unifying and generalizing these uses of commutativity under the abstraction afforded by ADTs.

@@

Though *commutativity* is often discussed in terms of an operation commuting with all other operations, it is actually more nuanced. If a pair of operations commute, then executing them in either order will produce the same result. Using the definitions from \cite{Kulkarni:PLDI11}, whether or not a pair of method invocations commute is a function of the methods, their arguments, their return values, and the *abstract* state of their target. We call the full set of commutativity rules for an ADT its *commutativity specification.* An example specification for a *Set* is shown in \autoref{spec}. It is important to note that interface choice affects commutativity: if `add(x)` returned a boolean that expressed if the item was added or not, then `add(x)` would only commute with `add(y)` if $x \ne y$. In this case, the difference doesn't drastically impact commutativity, but it would affect the cost of checking commutativity dynamically. There are actually many valid specifications which expose less than the maximum commutativity, but may be cheaper to implement.\cite{Kulkarni:PLDI11} ADT designers can navigate these choices to optimize an implementation.

@@

Sometimes, due to heavily skewed workloads such as those arising from social networks, there may be so many requests to a single record that just executing all the operations sequentially is a bottleneck which prevents scaling. In addition to transactional boosting, which we evaluate in this work, there are many other uses for commutativity. We mention just a few here which we have yet to evaluate.

**Record splitting.**
Associativity, often paired with commutativity, can be leveraged to execute operations in parallel on a "split" record and the changes made to these splits can be merged to form a consistent whole. Doppel \cite{Narula:OSDI14} used this for a handful of operations. We propose generalizing this concept to ADTs: associative operations execute on replicas first, other operations wait to execute until the replicas have been brought back in sync.

**Combining.**
Another spin on associative operations is to merge or *combine* operations as they come in. This is known as combining \cite{flatCombining,yew:combining-trees,funnels}, and it can drastically reduce contention. Combining can be done hierarchically: first with a few neighbors, then with clusters of neighbors, finally the combined operation is applied to the shared data structure.

# Evaluation {#eval}
To demonstrate the efficacy of leveraging commutative operations in transactions, we built a simple prototype key-value store, modeled after Redis, that supports complex data types for records, each with their own set of operations. 
The transaction protocol employs standard two-phase commit, and two-phase locking; more details are in the appendix (\autoref{apx:protocol}).
